Cultural Evolution of Conservation Behavior at the Intersection of Working

### 2020-10-02 DissertationProposal/DissertationProposal.Rmd
model desc

### Model Description

```{r conceptDiagram, echo=FALSE, fig.cap="Concept diagram of agent-based model scheduling under single group dynamics", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/ConceptDiagram.png")
```

## Preliminary Work

@@

```{r stovePic, echo=FALSE, fig.cap="Example of an improved, fuel efficient cookstove on Pemba Island, Tanzania. Image from Same Mwanga Environmental Conservation Advisory Organization", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/cookstove.jpg")
```

@@

```{r mapp, echo=FALSE, fig.cap="Satellite map of Pemba Island, Tanzania. Three marked locations are the wards which will be surveyed for fuel-efficient cookstove copying fidelity and outcomes.", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/BadPembaMap.png")
```

@@

```{r csp2, echo=FALSE, fig.cap="Simulated outputs showing a possible relationship between stove copying fidelity and geographic distance between the demonstrator and copier", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/CSP2.png")
```
# Conclusion
[@dudley2018essential]

>>>>

### Model Description
In the most simple scenario, the agent based model represents a simplistic world where there is one homogeneous group of agents who harvest resources from a landscape. This landscape is divided into two types, working and protected. Resources move between these patches at each time step with some probability. At time = 0, the protected and working landscapes are both at an equal, designated percent carrying capacity. At each time step, agents aim to maximize their harvest up until a designated harvest max. At time = 0, agents are assigned as either obligatory cooperators or as having the ability to defect to a small degree, spending a small portion of their time harvesting from the protected landscape.

The model also allows for ecological variation during the initial specification. The variable ecological starting conditions are as follows: Resource mobility designates the probability that a resource unit will move from the protected landscape into the working landscape or vice versa at each time step. Protected area size demarcates what percentage of the available resource units are within the protected area. Percent carrying capacity sets the total number of harvestable resource units as a percent of the landscape carrying capacity. Lastly, the resource regeneration rate designates what percentage the resource regenerates at each time step.

The environment also responds to agent behavior. At every time step, both the protected and working landscapes lose exactly the number of resources that agents harvest from them. The areas then regenerate resources according to the regeneration rate per time step and lastly, the resources move from one area to the other with the probability specified during model initiation.

Agents behave according to a specific set of rules. Agents “choose” to spend a given percent of their time harvesting from the protected area (defecting) and a percent of their time harvesting from the working landscape (defecting). Full defection is when the average percent of time agents spend in the protected area is equal to the percent of the landscape protected (i.e. they do not treat the protected area any differently than the working landscape). Agents learn these behaviors according to their specific learning mechanism. The learning will be an adjustable scale which assigns either a conformist-biased or success-biased learning strategy to a percentage of the agents. Behaviors may be modified at each time step by the key processes described below.

For the given number of timesteps, the model follows a set of key processes. At time step one, all agents harvest first spend their initial assigned percentage of effort harvesting from the working landscape. If agents reach harvest max, they do not harvest from the protected landscape, even if they are assigned as having the ability to defect. If they do not reach harvest max, they will spend the rest of their effort harvesting from the protected area. They will harvest up to the harvest max if possible. The resulting resources in both landscapes are multiplied by the regeneration factor, not past the carrying capacity of the landscape. Resources move between the protected and working landscapes with a probability proportional to designated probability of mobility. If agents are following a success biased learning strategy, they are paired with another agent at random and compare strategies. If one agent preformed worse than another in the first time step, the underperforming agent will adopt a strategy in-between (split the time difference in half) their previous strategy and the strategy of the agent who outperformed them. If agents are following a conformist strategy, they are grouped with 5 other agents at random at each time step, and adopt the mean strategy of all agents in their group. This process repeats for all time steps. This process is summarized conceptually in Fig \@ref(fig:conceptDiagram).

In the multiple competing groups scenario, two groups with separate resource pools will follow the processes described above. However, at the end of each time step the groups will compare the average harvest rate each group. Whichever group has the highest average harvest rate for that timestep will report the average harvest times for the protected and working landscapes. A specified percent of the agents from the less successful group will subsequently adopt the average harvest strategy of the more successful group.

In both single and competing group scenarios, the dependent variables of interest are as follows.: Cooperative behavior reports the average time that agents spend harvesting from the working landscape in each time step. Full defection is defined as all agents harvesting from the protected landscape for a proportion of time equal to the percent of the resource which is protected. Protected area degradation measures the percent carrying capacity of the protected landscape. Working landscape degradation measures the percent carrying capacity of the working landscape. Lastly, mean payoff to individuals measures the average payoff that agents receive at each time step.

```{r conceptDiagram, echo=FALSE, fig.cap="Concept diagram of agent-based model scheduling under single group dynamics", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch1/ConceptDiagram.png")
```

## Preliminary Work
I have developed a model which follows the above processes for one group with one resource pool. An interactive application which performs one run of that model can is found [at this link](https://mattclarknewaccount.shinyapps.io/ABM_Output_trial/).

This model currently assigns one learning strategy to all agents in the population. Users can select either successes-biased or conformity-biased learning to be applied. Future development will allow users to specify the percent of the population employing each learning strategy as specified in the methods above.

Environmental conditions can be set to provide neither an over abundance of resources nor a restrictive scarcity. Under these conditions, we see conformity biases maintain starting group norms around harvest behaviors. Of greater interest is that under success biased learning, we observe that protected areas enhance agent wellbeing (average harvest) initially, but defection eventually spreads and causes a failure of the protected area. With this in-mind, the next step for this research is to introduce a second group dynamic which will allow us to test the logic of cultural group selection for costly conservation behaviors under these conditions. Example outputs from this model can be seen in Fig \@ref(fig:abmOutput).

@@

```{r stovePic, echo=FALSE, fig.cap="Example of an improved, fuel efficient cookstove on Pemba Island, Tanzania. Image from Same Mwanga Environmental Conservation Advisory Organization", out.width = "50%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/cookstove.jpg")
```

@@

```{r mapp, echo=FALSE, fig.cap="Satellite map of Pemba Island, Tanzania. Three marked locations are the wards which will be surveyed for fuel-efficient cookstove copying fidelity and outcomes.", out.width = "75%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/BadPembaMap.png")
```

@@

```{r csp2, echo=FALSE, fig.cap="Simulated outputs showing a possible relationship between stove copying fidelity and geographic distance between the demonstrator and copier", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/CSP2.png")
```
# Conclusion
This research seeks to make clear how and why conservation behaviors succeed or fail to spread efficiently within a population. Moreover, the aim of this dissertation is to provide generalizable insights which can be applied to conservation initiatives in a variety of environments and population structures. A key advance made by this work is to link the fields of conservation social science and cultural group selection. Cultural group selection theory makes important predictions about how cooperative behaviors like conservation adoption may emerge and spread within populations, but has not yet been tested in a conservation context explicitly. Similarly, conservation social science incorporates environmental phenomena like resource mobility and regeneration, but largely omits generalizable insights about human behavioral evolution. By linking these two fields, we are greatly increasing the applicability of both bodies of work.

This dissertation will achieve this goal in three progressive steps. The first will formalize theory around cultural group selection in conservation contexts. This step is critical as conservation is a fundamentally different system from where these theories were originally developed. The second step will empirically validate these theoretical insights by comparing calculated cultural group selection pressures to real-world adoption of Community Forest Management Agreements on Pemba Island, Tanzania. Lastly, this project will make use of a conservation initiative currently spreading through Pemban communities, to specifically examine how behavioral efficiency may be altered by social contexts (e.g. communication frequency, genetic relatedness) as a behavior spreads throughout a population. All together, this research will allow us to better understand and possibly predict the emergence and effective spread of conservation norms within a population.

Human wellbeing is definitively tied to the persistence of renewable resources and ecosystem services. We are however, degrading these resources more rapidly than they can regenerate; reducing the stock of available resources while notably also slowing further regeneration. Conservation scientists have identified focal areas which, if preserved, may provide a disproportionate benefit to ecosystem function and offer a lifeline for human wellbeing. Simply put, we understand the environmental targets that we must achieve to ensure continued availability of renewable resources and we understand how conservation can achieve those targets. What we do not understand is how to effectively implement these required conservation measures. Reaching global biodiversity goals such as those set by the UN Convention on Biological Diversity will require “big and bold” conservation initiatives to be adopted at every level of the human population [@convention2010strategic; @dudley2018essential]. Currently however, even small-scale, local conservation measures often fail to gain traction and effectively run aground. This research will provide crucial theory and empirical evidence for how we can better shape and implement effective conservation initiatives at a variety of scales and in a variety of environmental conditions.
